Maximum Entropy Spectral Analysis: a case study

### 2021-02-19 paper/mesa.tex

This section is devoted to an extensive study on how to make such choice.
First of all (\ref{sec:optimizers}), we are going to define three different ``metrics" to measure how well the algorithm is performed: we will refer to them as \textit{optimizers}. The basic idea is to monitor, as the autoregressive order considered increase, the performance of the algorithm (in some sense) and pick, among the orders the one that yields better results.
Second, the performance of different optimizers will be assessed by performing the spectral analysis on different time series: a simple white gaussian noise (\ref{sec:gaussian_noise_validation}) and a more complicate PSD (\ref{sec:arp_validation}).
In a last subsection~\ref{sec:LIGO_validation} real data from LIGO-Virgo gravitational waves observatory.

\sschmidt{Qui ho provato a mettere un po' di ordine alla sezione. Ho scritto l'introduzione sopra per spiegare i passaggi logici della validation. Spero che vada bene. La parte sotto è tutta da rivedere e probabilmente c'è molto da asciugare: sono quasi 10 pagine! Mi metto a farlo subito}

@@

\item \textbf{Optimum Bayes Decision rule (OBD)} 
The last criterion we will consider is the OBD (Rao, 1982)\cite{doi:10.1029/WR018i004p01097}. With the technique of asymptotic integration, th optimizer is proposed to be the length that maximizes the posterior distribution \sschmidt{Which posterior?}: 
   \begin{equation}
        OBD(m) = (N - m - 2) \log(P_m) + m \log(N) + \sum_{k = 0}^{m-1} \log(P_k) + \sum_{k = 1}^{m} a_k^2.
    \end{equation}
\end{itemize}

@@

\subsection{Choosing an optimizer - Gaussian Noise} \label{sec:gaussian_noise_validation}
To apply Burg's method, it is a fundamental task to understand what is the best choice for estimating the order for the autoregressive stochastic process. Before applying it to real datasets, we will study main properties of the three different optimizer, introduce in section \ref{sec:optimizers} on distinct, simulated, realizations of some AR(p) processes. We will generate different Gauss noise time series with a given known power spectral density (PSD) to understand how each method performs in different situation. The analysis will concern the quality, to be defined later, of the reconstruction of the spectrum itself. We will consider two different kinds of spectra: i) a Gaussian spectrum with a given mean and variance; ii) an Advanced LIGO design sensitivity theoretical spectral curve.  As a part of our investigation, we will consider both ensemble-averaged, indicated with $\bar S$ and non averaged quantities. \\ 
In either cases, we will quantify the agreement between the reconstructed quantities and the expected ones.\\
A short description of the principle of the algorithm used to construct the time series is also given.
\subsection{Choosing an optimizer - AR(p) Time Series} \label{sec:arp_validation}
We will apply Burg method with each optimizer on a large amount of simulated dataset, obtaining a variety of estimate for PSD. We will study statistical properties for both single reproductions and ensemble average. In many cases, such as gravitational waves, one cannot observe multiple time series realizations for a fixed process, so that single reproduction studies are probably more useful. Still, study of ensemble average is important to highlight some interesting properties. In this way we can fully characterize the performance of each optimizer .
        \includegraphics[width = \linewidth]{Images/NormalPSD/NormalNoise.pdf}
    \includegraphics[width = \linewidth]{Images/NormalPSD/FPEpsdAndResiduals.pdf}

@@

\subsection{Analysis of Ligo Spectrum} \label{sec:LIGO_validation}
\begin{figure}
    \centering
    \includegraphics[width = \linewidth]{Images/LIGOsimulate/LigoPSDoriginal.pdf}
    \caption{Original power spectral density from Ligo}
    \label{fig:LigoPSDOriginal}
\end{figure}
In this subsection we concentrate again on the reconstruction of a specific, known, power spectrum.
We analyze the properties of reconstruction of this second spectrum whose shape is the Advanced LIGO design sensitivity theoretical spectral curve (Figure \ref{fig:LigoPSDOriginal}). 
For this analysis we will study properties only for arrays of 40960 points. We consider a sampling rate of $2048 samples / second$ that means we can reconstruct a Nyquist frequency of $1024 Hz$. We chose a longer dataset than before because algorithm works with a 1 dimensional interpolation of spectrum, and with shorter datasets the interpolated spectrum is not consistent with the original spectrum and does not capture the height of the peaks. Shorter choices would result in noise extracted from a different power spectrum than the one in figure \ref{fig:LigoPSDOriginal}. \\ 
We will study statistical properties from samples obtained with 500 simulations and compare the obtained results for each optimizer. 
We start considering statistical properties of mean spectrum obtained averaging over all the simulations and conclude with an analysis for the statistics of the reproduction of single spectrum. 
\subsubsection{Properties for the average spectrum}
Here we report mean properties of the reproductions of the spectrum. The spectrum we want to reconstruct is reported in Figure \ref{fig:ligospectrum},
\begin{figure}
    \centering

>>>>

This section is devoted to an extensive study on how to make such choice.
First of all (\ref{sec:optimizers}), we are going to define three different ``metrics" to measure how well the algorithm is performed: we will refer to them as \textit{optimizers}. The basic idea is to monitor, as the autoregressive order considered increase, the performance of the algorithm (in some sense) and pick, among the orders the one that yields better results.
Second, the performance of different optimizers will be assessed by performing the spectral analysis on time series with given analytical PSD, sec.~\ref{sec:arp_validation}.
In a last subsection~\ref{sec:LIGO_validation} real data from LIGO-Virgo gravitational waves observatory.

\sschmidt{Qui ho provato a mettere un po' di ordine alla sezione. Ho scritto l'introduzione sopra per spiegare i passaggi logici della validation. Spero che vada bene. La parte sotto è ancora da rivedere e probabilmente c'è molto da asciugare: sono quasi 10 pagine!}

@@

\item \textbf{Optimum Bayes Decision rule (OBD)} 
The last criterion we will consider is the OBD (Rao, 1982)\cite{doi:10.1029/WR018i004p01097}. With the technique of asymptotic integration, th optimizer is proposed to be the length that maximizes the posterior distribution \sschmidt{Which posterior?}: 
   \begin{align}
        OBD&(m) = (N - m - 2) \log(P_m) \nonumber\\
        &+ m \log(N) + \sum_{k = 0}^{m-1} \log(P_k) + \sum_{k = 1}^{m} a_k^2.
    \end{align}
\end{itemize}

@@

\subsection{Choosing an optimizer - A gaussian PSD} \label{sec:arp_validation}
We test the performance of the three optimizers on a random time series with a known power spectral density.
We take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$, see fig.~\ref{fig:autocorr}.
Of course, the time series will be generated by sampling from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$ and performing the inverse Fast Fourier Transform on the frequency series.

We will apply Burg method with each optimizer on a number of simulated time series, obtaining a variety of estimate for PSD.
We will study statistical properties for both single reproductions and ensemble average. In either cases, we will quantify the agreement between the reconstructed quantities and the expected ones.
In many realistic cases, such ensemble average is not accessible (one car rarely observe multiple time series realizations for the \textit{same} process), so that single reproduction studies are probably more useful. Still, study of ensemble average is important to highlight some interesting properties. In this way we can fully characterize the performance of each optimizer .
n what follows, we will indicate with $\bar S$ any ensemble averaged quantity.

We will attentively study the role of the filter length estimate M in determining the accuracy of the reconstruction. We will consider every optimizer separately and compare result only at the end of this section.
In section~\ref{sec:LIGO_validation}, we will perform the analysis again using, for the purpose of generating the time series, the Advanced LIGO design sensitivity PSD \cite{Ligo}.

\begin{figure}
	\centering
	\includegraphics[width = \linewidth]{Images/Noise and PSD/NormalPSDautocorr.pdf}
	\caption{A priori spectrum (left) and autocorrelation (right)}
	\label{fig:autocorr}
\end{figure}

\begin{figure}
    \centering
        \includegraphics[width = \linewidth]{Images/NormalPSD/NormalNoise.pdf}
    \caption{Noise in both time and frequency domain. Frequency noise behaviour is representative of the chosen spectrum, having its maximum amplitude around $f = 2.5 au$}
    \label{fig:noise}
\end{figure}

For our studies, we generate a dataset of 1000 time series of 3000 points each: see fig.~\ref{fig:noise} for an example of a time series in frequency and time domain.
For every time series we will compute the spectrum with every optimization method.

\sschmidt{Sono arrivato fino a qui nelle modifiche}

\subsubsection{Final Prediction Error (FPE)}
The estimate of the recursive order obtained with FPE on the 1000 realization of noise, is found to lay between 30 and 70. Qualitatively, there is a good agreement between the reconstructed spectrum and the true spectrum: a logarithmic plot of the mean of all the spectra versus the true spectrum is reported in \textbf{Figure \ref{fig:FPEmean}} with 90\% credible regions. It is evident that, despite not too large error bars, the tail of the normal distribution is not comprised in the interval of the estimation. 
\begin{figure}[H]
    \centering
    \includegraphics[width = \linewidth]{Images/NormalPSD/FPEpsdAndResiduals.pdf}
    \caption{Ensemble Average for FPE reconstruction with 90\% Credibility Regions (above) and percentage error (below)}
    \label{fig:FPEmean}
\end{figure}
To assess the magnitude of the disagreement, we compute the relative error between the mean spectrum $\hat{S}$ and the true spectrum $S$ for every frequency bin
\begin{equation}
    \nonumber
    r(f) = \frac{|\hat{S}(f) - S(f)|}{S(f)}\,.
\end{equation}
We found this to be a better choice than residuals since relative errors are not dependent on the magnitude of the data. Second Fig \ref{fig:FPEmean} (below) shows that the largest error in the reconstruction of the spectrum is in the tails, while in the bulk of the distribution we have a good approximation of the true function. 
For the entire spectrum, the mean relative error $\Bar{r}$ is 
\begin{equation}
    \nonumber
    \bar{r}_{\bar S, FPE} = 0.025.
\end{equation}
FPE performs well in reconstructing the spectrum even with relatively short time series, with a deviation from the true spectrum around 2\%, but it is evident that it is not able to capture all the features of the spectrum, since the reconstruction in the tails is inaccurate. If we compute the average percentage error in three standard deviations from the mean we gain an order of magnitude estimate of the accuracy of the reconstruction, obtaining: 
\begin{equation}
    \nonumber
    \bar{r}_{\bar S, FPE}^{\mu \pm 3\sigma} = 0.005. 
\end{equation}
and we can state the the method is highly performant in this frequency window. 
Since it is not always possible to work with more then one realization for a process, as in observational experiments, it is important to understand how this method behaves when we apply it on a single time series. In that case, the mean error and its standard deviation are: 
\begin{equation}
    \nonumber
    \bar r_{FPE} = 0.16; \quad \sigma_{r} = 0.16\,.
\end{equation}
The average error of a reconstruction based on a single time series is bigger than the one obtained considering the average of all the reconstructions.
Standard deviation of the errors assume small values and 90\% credibility regions (Cr.) in \ref{fig:FPEmean} are tight. This means that if we randomly take two of the reconstructed spectra, we expect their differences to be statistically small. These facts are evidence for FPE to be a relatively stable optimizer. 
We already said that the estimate for filter's length lays in a well defined range of values. The quality of the reconstruction, as noted in Chapter 1, depends on the order of the reconstructed AR process. 
To investigate this idea further, we plot the value of the optimizer for each recursion and the residuals as a function of the estimate for the order of the autoregressive process (\textbf{Figure \ref{fig:FPEErrorOrder}}),

@@

\subsection{Choosing an optimizer - LIGO Spectrum} \label{sec:LIGO_validation}
\begin{figure}
    \centering
    \includegraphics[width = \linewidth]{Images/LIGOsimulate/LigoPSDoriginal.pdf}
    \caption{Original power spectral density from Ligo}
    \label{fig:LigoPSDOriginal}
\end{figure}
In this subsection we concentrate again on the reconstruction of a specific, known, power spectrum.
We analyze the properties of reconstruction of this second spectrum whose shape is the Advanced LIGO design sensitivity theoretical spectral curve (Figure \ref{fig:LigoPSDOriginal}). 
For this analysis we will study properties only for arrays of 40960 points. We consider a sampling rate of $2048 samples / second$ that means we can reconstruct a Nyquist frequency of $1024 Hz$. We chose a longer dataset than before because algorithm works with a 1 dimensional interpolation of spectrum, and with shorter datasets the interpolated spectrum is not consistent with the original spectrum and does not capture the height of the peaks. Shorter choices would result in noise extracted from a different power spectrum than the one in figure \ref{fig:LigoPSDOriginal}. \\ 
We will study statistical properties from samples obtained with 500 simulations and compare the obtained results for each optimizer. 
We start considering statistical properties of mean spectrum obtained averaging over all the simulations and conclude with an analysis for the statistics of the reproduction of single spectrum. 
\subsubsection{Properties for the average spectrum}
Here we report mean properties of the reproductions of the spectrum. The spectrum we want to reconstruct is reported in Figure \ref{fig:ligospectrum},
\begin{figure}
    \centering
